---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pairedGSEA

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![Codecov test coverage](https://codecov.io/gh/shdam/pairedGSEA/branch/master/graph/badge.svg)](https://app.codecov.io/gh/shdam/pairedGSEA?branch=master)
<!-- badges: end -->



`pairedGSEA` is an R package that helps you to run a paired differential gene expression (DGE) and splicing (DGS) analysis.
Providing a count matrix, `pairedGSEA` combines the results of `DESeq2` (DGE) and `DEXSeq` (DGS), aggregates the p-values to gene level, and allows you to run a subsequent gene set over-representation analysis using its implementation of the `fgsea::fora` function. 


## Article

A preprint on pairedGSEA is available [here](https://doi.org/10.1101/2022.08.29.505720).


## Installation


``` {r install, eval=FALSE}
# Install Bioconductor packages with:
# devtools::install_bioc(c("DESeq2", "DEXSeq", "fgsea", "SummarizedExperiment", "S4Vectors", "sva", "BiocParallel"))

# Install pairedGSEA
devtools::install_github("shdam/pairedGSEA")

```



## Documentation

To view documentation for the version of this package installed in your system, start R and enter:

```{r vignettes, eval=FALSE}
browseVignettes("pairedGSEA")
```




## Usage

Please see the User Guide vignette for a detailed description of usage.

Here is a quick runthrough of the functions:


<br>
Load example data.
```{r data}
library("pairedGSEA")
suppressPackageStartupMessages(library("SummarizedExperiment"))

data("example_se")
example_se
```

Run paired differential analysis

```{r diff}
set.seed(500) # For reproducible results

diff_results <- paired_diff(
  example_se,
  group_col = "group_nr",
  sample_col = "id",
  baseline = 1,
  case = 2,
  store_results = FALSE,
  quiet = TRUE
  )
diff_results
```

Over-representation analysis of results

```{r ora}
# Define gene sets in your preferred way
gene_sets <- pairedGSEA::prepare_msigdb(species = "Homo sapiens", category = "C5", gene_id_type = "ensembl_gene")

ora <- paired_ora(
  paired_diff_result = diff_results,
  gene_sets = gene_sets
  )
ora
```

You can now plot the enrichment scores against each other and identify pathways of interest.

```{r plot}
plot_ora(ora)
```


## Report issues

If you have any issues or questions regarding the use of pairedGSEA, please do not hesitate to raise an issue on GitHub. In this way, others may also benefit from the answers and discussions.
