---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pairedGSEA

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/shdam/pairedGSEA/branch/master/graph/badge.svg)](https://app.codecov.io/gh/shdam/pairedGSEA?branch=master)
<!-- badges: end -->



`pairedGSEA` is an R package that helps you to run a paired differential gene expression (DGE) and splicing (DGS) analysis.
Providing a bulk RNA count data, `pairedGSEA` combines the results of `DESeq2` (DGE) and `DEXSeq` (DGS), aggregates the p-values to gene level, and allows you to run a subsequent gene set over-representation analysis using its implementation of the `fgsea::fora` function. 


## Article

`pairedGSEA` is published in [BMC Biology](https://doi.org/10.1186/s12915-023-01724-w).

Please cite with `citation("pairedGSEA")`

## Installation

Dependencies 
``` {r install_dep, eval=FALSE}
# Install Bioconductor dependencies
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("SummarizedExperiment", "S4Vectors", "DESeq2", "DEXSeq", "fgsea", "sva", "BiocParallel"))
```

Install `pairedGSEA` from Bioconductor
```{r install_bioc, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("pairedGSEA")
```

Install development version from GitHub
``` {r install_github, eval=FALSE}
# Install pairedGSEA from github
devtools::install_github("shdam/pairedGSEA", build_vignettes = TRUE)
```

## Documentation

To view documentation for the version of this package installed in your system, start R and enter:

```{r vignettes, eval=FALSE}
browseVignettes("pairedGSEA")
```



## Usage

Please see the User Guide vignette for a detailed description of usage.

Here is a quick run-through of the functions:


<br>
Load example data.
```{r data}
suppressPackageStartupMessages(library("SummarizedExperiment"))
library("pairedGSEA")

data("example_se")
example_se
```

Run paired differential analysis

```{r diff}
set.seed(500) # For reproducible results

diff_results <- paired_diff(
  example_se,
  group_col = "group_nr",
  sample_col = "id",
  baseline = 1,
  case = 2,
  store_results = FALSE,
  quiet = TRUE
  )
```

Over-representation analysis of results

```{r ora}
# Define gene sets in your preferred way
gene_sets <- pairedGSEA::prepare_msigdb(
    species = "Homo sapiens",
    category = "C5",
    gene_id_type = "ensembl_gene"
    )

ora <- paired_ora(
  paired_diff_result = diff_results,
  gene_sets = gene_sets
  )

```

You can now plot the enrichment scores against each other and identify pathways of interest.

```{r plot}
plot_ora(
    ora, 
    paired = TRUE # Available in version 1.1.0 and newer
    ) +
    ggplot2::theme_classic()
```


## Report issues

If you have any issues or questions regarding the use of `pairedGSEA`, please do not hesitate to raise an issue on GitHub. In this way, others may also benefit from the answers and discussions.
